Deciding Bitvector Termination with Program Synthesis

### 2014-06-27 papers/termination/paper.tex
sec 6 stuff

\section{Termination as Second-Order Satisfaction} \label{sec:second.order}
In this section we will show how several variations of the halting problem can be defined in second-order
logic.  In fact, many of these problems can be encoded in a fragment of second-order logic we will
suggestively refer to as the \emph{synthesis fragment}:

@@

\begin{figure*}
\begin{framed}
\begin{definition}[Second-order Termination Formula]
\label{def:termination-formula}
\begin{align*}
 \exists R . \forall x, x' . & L(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
\end{align*}
\end{definition}

\begin{definition}[Conditional Termination Formula]
\label{def:conditional-termination-formula}
 \begin{align*}
  \exists R, V . \forall x, x' . & I(x) \wedge L(x) \rightarrow V(x) ~ \wedge \\
                                 & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

@@

The existence of a ranking function for a single loop is equivalent to the satisfiability
of the formula from Definition~\ref{def:termination-formula}.  If this formula is satisfiable,
$R$ is a ranking function that proves loop $L$ terminates.

@@

Non-termination of the loop $L$ is encoded by Definition~\ref{def:nonterm-formula}.
If this formula is satisfiable, $N$ is a recurrence set for $L$ which proves
$L$'s non-termination.  However this formula has one more level of quantifier alternation
than the others (it is an $\exists \forall \exists$ formula), which will cause us problems
shortly.  We observe that in the case that $T$ is deterministic (that is, each state has
exactly one successor),  the non-termination problem is equivalent to the
satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.

>>>>

\section{Termination as Second-Order Satisfaction} \label{sec:second.order}
In this section we will show how several variations of the halting problem can be defined in second-order
logic.  
\begin{lstlisting}
while (L) { T }
\end{lstlisting}

 In fact, many of these problems can be encoded in a fragment of second-order logic we will
suggestively refer to as the \emph{synthesis fragment}:

@@

\begin{figure*}
\begin{framed}
\begin{definition}[Second-order Termination Formula]
\label{def:termination-formula}
\begin{align*}
 \exists R . \forall x, x' . & G(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
\end{align*}
\end{definition}

\begin{definition}[Conditional Termination Formula]
\label{def:conditional-termination-formula}
 \begin{align*}
  \exists R, V . \forall x, x' . & I(x) \wedge G(x) \rightarrow V(x) ~ \wedge \\
                                 & G(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

@@

We first consider a single loop $L = (I,G,T)$ consisting of the initial states $I : X \rightarrow B$, the loop guard $G : X \rightarrow B $ and a body $T : X \rightarrow X$.
The existence of a ranking function for this loop is equivalent to the satisfiability
of the formula from Definition~\ref{def:termination-formula}.  If this formula is satisfiable,
$R$ is a ranking function that proves loop $L$ terminates.

@@

Gupta et al. \cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination of a transition relation $T$ by the existence of a
\emph{recurrence set} (open recurrence set in the terminology of \cite{DBLP:conf/tacas/ChenCFNO14}), 
i.e. a nonempty set of states $G$ such that for each $s \in G$ there
exists a transition to some $s'\in G$. 
For our case, given a loop $L = (I,G,T)$, the notion of open recurrent set is encoded by Definition~\ref{def:nonterm-formula}.
If this formula is satisfiable, $N$ is an open recurrence set for $L$ which proves
$L$'s non-termination. The issue with this formula is the additional level of quantifier alternation as compared to the synthesis fragment
(it is an $\exists \forall \exists$ formula). %This means that we will not be able to use the solver for the synthesis fragment to solve it.

In order to formulate the non-termination problem in the synthesis fragment, we make use of the notion of \emph{closed recurrence set} introduced by Chen et al. 
in \cite{DBLP:conf/tacas/ChenCFNO14}. As ooposed to a closed recurrence set, this is a  purely universal property: for each $s \in G$, each of its successors 
must be in the recurrence set. The non-termination problem thus is equivalent to the
satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.

\cite{DBLP:conf/popl/GuptaHMRX08,DBLP:conf/tacas/ChenCFNO14} make use of the syntactic notion of \emph{lasso}.
%The standard way of proving non-termination is by detecting non-terminating lassos \cite{DBLP:conf/popl/GuptaHMRX08,DBLP:conf/tacas/ChenCFNO14}.
A lasso at a cutpoint location consists of two sequences
of transitions, the stem and the loop.
The stem is a finite path from an initial location to the cutpoint location. 
The loop is a finite path that starts and ends at the cutpoint
location by following a cyclic path through the controlflow
graph. A lasso induces an infinite execution if the infinite path obtained by traversing the stem and then unrolling
the loop infinitely many times is feasible. In this case, the lasso is said to be non-terminating.
%% We observe that in the case that $T$ is deterministic (that is, each state has
%% exactly one successor),  the non-termination problem is equivalent to the
%% satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.

In contrast, our method is purely semantic ...
